---
title: "DS2020 Final Project Report"
author: "Hong Hao Khoo and Harsh Mishrikotkar"
date: "2025-04-28"
output: github_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rvest)
library(tidyverse)
library(ggplot2)
library(stringi)
library(ggpubr)
library(grid)
library(corrplot)
library(caret)
library(pROC)
library(readr)
library(stringr)
library(dplyr)
library(gridExtra)
```

# Benign or Malignant Tumor
## Introduction/Background
A tumor is a lump or mass of tissue that forms when cells grow and divide too much or don’t die when they should. The word tumor is often used interchangeably with cancer. Cancer is a group of diseases characterized by the uncontrolled growth and spread of abnormal cells in the body. Scientifically, tumors can be categorized into benign or malignant. Benign tumors do not invade nearby tissues or spread to other parts of the body, whereas malignant tumors do. Malignant tumors are what we call cancer. If it is malignant, then it is a cancer otherwise it is not.

After a patient is diagnosed with a tumor, a crucial next step is determining whether it is benign or malignant. This is often done through a biopsy, which can be invasive, painful, and stressful. The goal of this project is to explore whether measurable tumor characteristics can help increase the chances of identifying the tumor type through imaging techniques and improve the accuracy of that identification. By analyzing features such as size, shape, and texture, we aim to support the development of more accessible, non-invasive diagnostic tools that can complement or even reduce the reliance on biopsies.

## Project Goals and Questions
When diagnosed with tumor but not knowing if it’s benign or malignant is a confusing and frightening moment - you're stuck waiting, hoping for good news but scared of the worst. In this project, we aim to analyze a data set containing records of patients diagnosed with tumors. Each patient is assigned a unique ID and labeled with the tumor type—either benign or malignant. The data set also includes various visual characteristics of the tumors, each represented by a corresponding numerical value.

Through the analysis of this data set, we hope to find out:

1. Is there a correlation between the characteristics and the type of tumor?

2. Which characteristic is the most highly correlated and which is lowest to the type of tumor?

3. Can we predict whether a tumor is benign or malignant base on its characteristics?

4. Is it possible to speed up the process of identifying whether a tumor is benign or malignant? 

5. Can we increase chances of identifying tumor type from imaging techniques?

6. Can we increase accuracy of identifying tumor type from imaging techniques?

## Data set
The link to the data set is https://www.kaggle.com/datasets/erdemtaha/cancer-data. The website contains a downloadable csv file, which is the tumor dataset. This dataset have 569 observations and 32 variables. The data set contain records of patients diagnosed with tumor. Each patient is given an unique ID along with the type of tumor, benign or malignant. It also contains the visual characteristics of the tumor and a value for each of these characteristics.  
```{r}
# Import dataset
data <- read_csv("Cancer_Data.csv")
data <- data[1:(length(data)-1)]

# View dataset
head(data, 10)
```

## Cleaning & Observation
```{r}
summary(data)

str(data)

# Convert diagnosis to factor
dataClean <- data %>%
  mutate(diagnosis = factor(diagnosis, levels = c("B", "M"), labels = c("Benign", "Malignant")))

# Drop ID column
dataClean <- dataClean %>% select(-id)

# Check for any missing values in the rest of the data set
colSums(is.na(dataClean))
```

For cleaning of the data, we assigned the cleaned version of our data set to a new data set called dataClean. For dataClean, we converted variable diagnosis to type factor with levels "B" and "M" and labels "Benign" and "Malignant". Secondly, the id column was dropped as it is not relevant. After that, we check if the data set have any missing values and there was none. 

## Variables
* diagnosis                 : Benign or Malignant cancer
* radius_mean               : Radius mean value
* texture_mean              : Texture mean value
* perimeter_mean            : Perimeter mean value
* area_mean                 : Area mean value
* smoothness_mean           : Smoothness mean value
* compactness_mean          : Compactness mean value
* concavity_mean            : Concavity mean value
* concave points_mean       : Concave points mean value
* symmetry_mean             : Symmetry mean value
* fractal_dimension_mean    : Fractal dimension mean value
* radius_se                 : Radius standard error value
* texture_se                : Texture standard error value
* perimeter_se              : Perimeter standard error value
* area_se                   : Area standarad error value
* smoothness_se             : Smoothness standard error value
* compactness_se            : Compactness standard error value
* concavity_se              : Concavity standard error value
* concave points_se         : Concave points standard error value
* symmetry_se               : Symmetry standard error value
* fractal_dimension_se      : Fractal dimension standard error value
* radius_worst              : Radius worst value
* texture_worst             : Texture worst value
* perimeter_worst           : Perimeter worst value
* area_worst                : Area worst value
* smoothness_worst          : Smoothness worst value
* compactness_worst         : Compactness worst value
* concavity_worst           : Concavity worst value
* concave points_worst      : Concave points worst value
* symmetry_worst            : Symmetry worst value
* fractal_dimension_worst   : Fractal dimension worst value

## Statistics Summary
```{r}
# Basic statistics summary for each column
stats_summary <- data.frame(
  Column = character(),
  Min = numeric(),
  Q1 = numeric(),
  Median = numeric(),
  Mean = numeric(),
  Q3 = numeric(),
  Max = numeric(),
  SD = numeric(),
  stringsAsFactors = FALSE
)

for (col_name in names(dataClean)[sapply(dataClean, is.numeric)]) {
  col_data <- dataClean[[col_name]]
  stats_summary <- rbind(stats_summary, data.frame(
    Column = col_name,
    Min = min(col_data, na.rm = TRUE),
    Q1 = quantile(col_data, 0.25, na.rm = TRUE),
    Median = median(col_data, na.rm = TRUE),
    Mean = mean(col_data, na.rm = TRUE),
    Q3 = quantile(col_data, 0.75, na.rm = TRUE),
    Max = max(col_data, na.rm = TRUE),
    SD = sd(col_data, na.rm = TRUE)
  ))
}

# Display summary statistics 
print(stats_summary)
```

These are the statistics summary for every numeric variable of the data set.

## Results
### Visual 1 (Tumor Type Distribution Bar Plot)
```{r}
# Show how many benign vs malignant cases there are
ggplot(dataClean, aes(x = diagnosis)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Tumor Types", x = "Diagnosis", y = "Count") +
  theme_minimal()
```

To understand the class distribution in our data set, we created a bar plot comparing the number of benign and malignant tumors. This visualization revealed that benign tumors are more common in the data set than malignant ones.

This class imbalance is important to note, as it may affect the performance of classification models. Models trained on imbalanced data might become biased toward predicting the majority class (benign), so additional techniques like resampling or adjusting evaluation metrics may be considered in future modeling steps.

### Visual 2 (Feature Correlation Heatmap)
```{r fig.width=10, fig.height=10}

# Convert diagnosis to numeric
dataClean$diagnosis_num <- ifelse(dataClean$diagnosis == "Malignant", 1, 0)

# Select numeric features (excluding the original diagnosis factor)
numeric_features <- select_if(dataClean, is.numeric)

# Compute correlation with diagnosis
cor_with_diag <- cor(numeric_features, use = "complete.obs")[, "diagnosis_num"]
cor_df <- data.frame(Feature = names(cor_with_diag), Correlation = cor_with_diag)

# Sort by absolute correlation
cor_df <- cor_df %>% arrange(desc(abs(Correlation)))

ggplot(cor_df, aes(x = reorder(Feature, Correlation), y = Correlation)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Correlation of Features with Diagnosis (Malignant = 1, Benign = 0)",
       x = "Feature", y = "Pearson Correlation") +
  theme_minimal()
```

The bar plot above visualizes the Pearson correlation between each numeric feature and the tumor diagnosis (where `0 = Benign` and `1 = Malignant`). Each bar represents a feature, and the length of the bar reflects how strongly that feature is linearly associated with the likelihood of a tumor being malignant. Features with higher positive correlation values are more likely to be found in malignant tumors, while features with near-zero correlation show little to no relationship with diagnosis.

From the plot, we observe that features such as `concave points_worst`, `perimeter_worst`, and `radius_worst` exhibit the strongest positive correlations with malignancy. This suggests that tumors with larger sizes and more irregular shapes (as captured by these features) are more likely to be malignant. In contrast, features like `smoothness_se`, `fractal_dimension_mean`, and `symmetry_se` have very low correlation values, indicating they are poor predictors for tumor classification in this dataset.

### Visual 3 (Histogram Overview Of All Variables In Relation To Diagnosis)
```{r fig.width=15, fig.height=45}
# Function to create a histogram with diagnosis coloring
create_histogram_by_diagnosis <- function(df, column_name) {
  # Calculate appropriate number of bins using Sturges' rule as a starting point
  # but capped between 10 and 30 bins
  n_distinct <- length(unique(df[[column_name]]))
  n_bins <- min(max(ceiling(log2(n_distinct) + 1), 10), 30)
  
  # Create histogram with diagnosis coloring
  p <- ggplot(df, aes(x = !!sym(column_name), fill = factor(diagnosis))) +
    geom_histogram(bins = n_bins, 
                   color = "black",
                   alpha = 0.7,
                   position = "identity") +
    scale_fill_manual(values = c("Benign" = "green", "Malignant" = "red"),
                      labels = c("0" = "Benign", "1" = "Malignant"),
                      name = "Diagnosis") +
    labs(title = column_name,
         x = column_name,
         y = "Frequency") +
    theme_minimal() +
    theme(plot.title = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(size = 8),
          axis.text.y = element_text(size = 8),
          axis.title = element_text(size = 9),
          legend.position = "bottom",
          legend.title = element_text(size = 9),
          legend.text = element_text(size = 8))
  
  return(p)
}

# Create histograms for all columns with diagnosis coloring
plot_list <- list()
column_names <- names(dataClean)

for (col_name in column_names) {
  # Skip the diagnosis column itself and any non-numeric columns
  if (col_name == "diagnosis" || !is.numeric(dataClean[[col_name]])) {
    cat("Skipping column:", col_name, "\n")
    next
  }
  
  # Create and store the histogram
  plot_list[[col_name]] <- create_histogram_by_diagnosis(dataClean, col_name)
}

# Arrange all histograms in a single grid with 2 columns
# Calculate the height needed based on number of plots
num_plots <- length(plot_list)
grid_height <- ceiling(num_plots/2) * 3

# Create a title for the grid
title <- textGrob("Histogram Distributions by Diagnosis", 
                 gp = gpar(fontsize = 14, fontface = "bold"))

# Arrange all plots in a grid with 2 columns
arranged_plots <- arrangeGrob(grobs = plot_list, 
                             ncol = 2,
                             top = title)

# Display the grid (may be very tall depending on number of features)
grid.newpage()
grid.draw(arranged_plots)
```

The histograms display the distribution of each numerical feature across benign and malignant tumor cases. From these visualizations, we observed that certain variables showed stronger separation between the two classes. In particular, `concave points_worst` showed the most distinct difference in value ranges between benign and malignant tumors, making it appear to be a strong predictor. However, in real-world diagnostic settings, relying on a `*_worst` measurement can be impractical. These values often require analyzing the three most extreme instances among many cell nuclei, which may not be feasible during early or routine examinations. 

### Visual 4 (Pairplot For *_mean Variables)
```{r fig.width=15, fig.height=15}
dataPairedPlot <- dataClean |>
  mutate(diagnosis = recode(diagnosis, "Benign" = 0, "Malignant" = 1))

pairs(dataPairedPlot[, 3:12], panel=function(x,y){
  # Get a vector of colors for each point in the plot
  colors <- ifelse(dataPairedPlot$diagnosis == 0, "green",
                   ifelse(dataPairedPlot$diagnosis == 1, "red", "blue"))

  # Plot the points with the corresponding colors
  points(x, y, col = colors)
})
```

We examined the pair plots of the `*_mean` variables, with each point colored based on diagnosis—green for benign and red for malignant. This visualization helped us observe how different features relate to each other and whether certain combinations of variables show a clear separation between the two diagnosis groups. By analyzing the clustering and spread of colored points, we could identify which variable pairs might be more useful in distinguishing malignant tumors from benign ones. This is useful for feature selection and understanding underlying patterns in the data.

### Visual 5 (Boxplot: Characteristics by Cancer Type)
```{r}
# Select all mean columns + diagnosis
mean_data <- dataClean %>%
  select(diagnosis, ends_with("_mean"))

# First plot: all mean features EXCEPT area_mean
mean_data %>%
  select(-area_mean) %>%
  pivot_longer(-diagnosis, names_to = "Feature", values_to = "Value") %>%
  ggplot(aes(x = Feature, y = Value, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Mean Characteristics (Excluding Area) by Diagnosis",
       x = "Feature", y = "Value") +
  scale_fill_manual(values = c("Benign" = "skyblue", "Malignant" = "salmon")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Second plot: only area_mean
ggplot(mean_data, aes(x = diagnosis, y = area_mean, fill = diagnosis)) +
  geom_boxplot() +
  labs(title = "Area Mean by Diagnosis", x = "Diagnosis", y = "Area Mean") +
  scale_fill_manual(values = c("Benign" = "skyblue", "Malignant" = "salmon")) +
  theme_minimal()
```

To explore which tumor characteristics differ between benign and malignant cases, we analyzed all `*_mean` features using side-by-side boxplots. The first plot compares all mean features except `area_mean`, which was plotted separately due to its significantly larger scale and would otherwise cause the remaining features to appear compressed and unreadable when plotted together.

From the visualizations, we observed that malignant tumors tend to have higher values for several key features, including `radius_mean`, `perimeter_mean`, `concavity_mean`, and `concave points_mean`. These features show a clear upward shift in the boxplots for malignant cases, suggesting that tumors with more aggressive or irregular growth patterns (such as larger size or more concave points) are more likely to be cancerous.

In contrast, several other features — including `compactness_mean`, `fractal_dimension_mean`, `smoothness_mean`, and `symmetry_mean` — appear tightly clustered near zero in the box plot. Their small numeric ranges make it difficult to visually distinguish between benign and malignant cases. While these features may still contain subtle patterns, their limited variability and overlap in values suggest they may not be as strong or reliable for classification without additional transformation or scaling. Simply speaking, these small-valued features may not provide enough contrast between benign and malignant tumors, making them weaker predictors on their own. They could still be useful when combined with other features, but by themselves they are not strong indicators of diagnosis.

The separate box plot for `area_mean` further reinforces this trend — malignant tumors have a much larger average area compared to benign ones. This aligns with biological expectations, as malignant tumors often grow more rapidly and invade surrounding tissue, resulting in a greater area.

These results support our research objective by identifying features that significantly differ between benign and malignant tumors. Such insights are crucial in building predictive models that can aid in early and accurate cancer diagnosis based on measurable tumor attributes.

### Visual 6 (Histogram: Top 5 Features)
```{r}
# List of top features to visualize
top_features <- c("radius_mean", "perimeter_mean", "area_mean", 
                  "concavity_mean", "concave points_mean")

# Loop through each feature and plot histogram
for (feature in top_features) {
  p <- ggplot(dataClean, aes(x = .data[[feature]], fill = diagnosis)) +
    geom_histogram(position = "identity", alpha = 0.6, bins = 30) +
    labs(title = paste("Distribution of", feature, "by Diagnosis"),
         x = feature, y = "Count") +
    scale_fill_manual(values = c("Benign" = "skyblue", "Malignant" = "salmon")) +
    theme_minimal() +
    theme(legend.position = "top")

  print(p)
}
```

The following set of histograms displays the distribution of five selected tumor characteristics (`radius_mean`, `perimeter_mean`, `area_mean`, `concavity_mean`, and `concave points_mean`) for benign and malignant tumors. Each plot shows the number of tumors (y-axis) falling within specific value ranges of a given feature (x-axis), with colors distinguishing the diagnosis: blue for benign and red for malignant.

These five features were chosen based on earlier boxplot analysis, where they demonstrated the most significant differences between benign and malignant groups.

The histograms reveal clear distribution shifts between benign and malignant tumors. For instance, the `radius_mean`, `perimeter_mean`, and `area_mean` histograms show that malignant tumors generally have larger values than benign ones, with minimal overlap. This aligns with biological expectations that malignant tumors tend to grow larger and invade surrounding tissue more aggressively.

Similarly, `concavity_mean` and `concave points_mean` — both indicators of irregular tumor shape — are skewed toward higher values for malignant tumors. Benign tumors, on the other hand, cluster near lower values for these features, suggesting smoother and less complex tumor boundaries.

These histograms visually reinforce the idea that malignant tumors are not only larger but also more structurally irregular, and that these characteristics can be used as key indicators in distinguishing tumor types. The distinct separation in their distributions provides valuable insight into the measurable differences that can aid in accurate diagnosis.

## Simple Logistric Regression Model
```{r}
# Split data
set.seed(42)
train_idx <- createDataPartition(dataClean$diagnosis_num, p = 0.8, list = FALSE)
train_data <- dataClean[train_idx, ]
test_data <- dataClean[-train_idx, ]

# Train logistic regression using top 5 features
log_model <- glm(diagnosis_num ~ radius_mean + perimeter_mean + area_mean + 
                   concavity_mean + `concave points_mean`, 
                 data = train_data, family = "binomial")
summary(log_model)

# Predict probabilities
pred_probs <- predict(log_model, newdata = test_data, type = "response")

# Convert probabilities to class labels
pred_class <- ifelse(pred_probs > 0.5, 1, 0)

# Confusion Matrix
confusionMatrix(factor(pred_class), factor(test_data$diagnosis_num), 
                positive = "1")

# ROC Curve and AUC
roc_obj <- roc(test_data$diagnosis_num, pred_probs)
plot(roc_obj, main = "ROC Curve for Logistic Regression")
auc(roc_obj)
```

To evaluate how well the selected tumor characteristics can classify cancer as benign or malignant, we trained a logistic regression model using the five most correlated features identified during EDA: `radius_mean`, `perimeter_mean`, `area_mean`, `concavity_mean`, and `concave points_mean`. These features were chosen based on earlier boxplots and correlation plots, where they showed clear separation between the two diagnosis groups.

The model achieved a high accuracy of 91.15%, with strong sensitivity (84.85%) and specificity (93.75%), indicating that it correctly identifies both malignant and benign cases in most situations. The ROC curve further supports the model's performance, with an AUC (Area Under the Curve) of 0.9833, which is considered excellent and demonstrates the model's strong discriminative power.

These results reinforce the potential of using imaging-derived features to predict cancer type, offering a less invasive and more accessible alternative to biopsy. While this model is simple, it shows that selected features like tumor size and shape irregularities are strong indicators and could support clinical decision-making when paired with other diagnostic tools.

## Conclusion
Through our exploration of tumor imaging data, we found several meaningful insights that help answer the original research questions.

**1. Is there a correlation between the characteristics and the type of tumor?**

Yes. Visual 2 (Feature Correlation Heatmap) clearly shows that several imaging-based characteristics are correlated with tumor type. Features such as concave points_worst, perimeter_worst, and radius_worst show strong positive correlations with malignancy, indicating that certain tumor traits are more common in malignant cases.

**2. Which characteristic is the most highly correlated and which is lowest to the type of tumor?**

Based on Visual 2, concave points_worst is the most highly correlated feature with malignancy. On the other hand, features like smoothness_se and texture_se have very low correlation values, making them poor indicators of tumor type.

**3. Can we predict whether a tumor is benign or malignant based on its characteristics?**

Yes. As demonstrated in Visual 7 (Simple Logistic Regression Model), using five selected features (radius_mean, perimeter_mean, area_mean, concavity_mean, and concave points_mean), the model achieved 91.15% accuracy with an AUC of 0.9833. This shows that these features can reliably be used to predict tumor type.

**4. Is it possible to speed up the process of identifying whether a tumor is benign or malignant?**

Yes. Visuals 5 and 6 (Boxplots and Histograms) show that some features provide immediate visual separation between benign and malignant tumors. These characteristics, when extracted from imaging, can support faster decision-making compared to traditional, time-consuming diagnostic procedures like biopsies.

**5. Can we increase chances of identifying tumor type from imaging techniques?**

Yes. Visuals 3, 5, and 6 show that imaging features like area_mean and concavity_mean allow for a clear distinction between tumor types. These strong visual patterns improve the chance of identifying tumor type through non-invasive imaging.

**6. Can we increase accuracy of identifying tumor type from imaging techniques?**

Yes. The logistic regression model in Visual 7 demonstrates that using a small set of well-chosen imaging features can yield highly accurate classification. This suggests that imaging can be used to reliably and accurately distinguish between benign and malignant tumors.

In summary, our analysis supports the use of imaging-derived characteristics, particularly those related to size and shape, for accurate and efficient tumor classification. These insights could aid in developing faster, more accessible diagnostic tools that improve early detection and clinical outcomes.